Statically Typed String Sanitation Inside a Python

### 2014-08-14 sanitation-psp14/sanitation-psp14-short.tex
Cleanup
Layout

We will begin in Sec. \ref{calculus} by specifying this type system minimally, as a conservative extension of the simply typed lambda calculus called $\lambda_{RC}$. This allows us to specify the guarantees that the type system provides precisely. We also formally specify a translation from this calculus to a typed calculus with only standard strings and regular expressions, intending it as a guide to language implementors interested in building this feature into their own languages. This also demonstrates that no additional space overhead is required.

@@

% TODO: need to update remainder of intro ... To demonstrate this approach, we present a simply typed lambda calculus with \emph{constrained strings}; that is, a set of string types parameterized by regular expressions. If $s : \stringin{r}$, then $s$ is a string matching the language $r$. Additionally, we include an operation $\rreplace{r}{s_1}{s_2}$ which corresponds to the replace mechanism available in most regular expression libraries; that is, any substring of $s_1$ matching $r$ is replaced with $s_2$. The type of this expression is the computed, and is likely ``smaller" or more constrained than the type of $s_1$. Libraries, frameworks or functions which construct and execute commands containing input can specify a safe subset $\stringin{r_{\sf spec}}$ of strings, and input sanitation algorithms can construct such a string using ${\sf rreplace}$ or, optionally, by coercion (in which case a runtime check is inserted). We also show how this system is translated into a host language containing a regular expression library such that the safety guarantee of the extended language is preserved.

% Summarily, we present a simple type system extension which ensures the absence of input sanitation vulnerabilities by statically checking input sanitation algorithms which use an underlying regular expression library. This approach is \emph{composable} in the sense that it is a conservative extension. This approach is also \emph{complementary} to existing input sanitation techniques which use string replacement for input sanitation.
% Our type system for regular string types, $\lambdas$, is defined as an exntension to a simply-typed lambda calculus.
% Operations on strings are given types according to corresponding, type-level
% computations on regular expressions. A concatenation operation is defined in 
% order to demonstrate this principle, but the premier operation of our system
% is \emph{replacement}, which captures a rich set of input sanitation techniques
% including escaped characters and encodings. In addition to defining this system,
% we present a correctness-preserving compilation of $\lambdas$ into an underlying
% language with ordinary strings as a regular expression library.

% In addition to this theoretical contribution, we also explain how our extension
% may be implemented in the Ace Programming Language, which is itself an extension
% of Python -- a popular web programming language.

% Summarily, we present a simple type system extension which ensures the absence
% of input sanitation vulnerabilities by statically checking the correctness and 
% correct use of input sanitation algorithms. This approach is \emph{composable}
% in the sense that it is a conservative extension. This approach is \emph{complementary}
% to existing input sanitation techniques that use regular expressions for input sanitation.

%%
% This is the paragraph which ties the paper back to the CFP.
%
% For refernce, here's a copy of the cfp:
%The importance of security and privacy in software engineering is now a pressing concern. 
%Over the last decade, many programmers have recognized the importance of including 
%security and privacy requirements at project start. This has led to efforts 
%in secure/security engineering and privacy engineering, which focus on guidelines and 
%best practices that can be used at the design stage to create safer code. 
%Unfortunately, these disciplines are not pervasive and still in their infancy.
%The goal of this workshop is to make security and privacy first class citizens in programming and programming languages.

%This forum aims to gather research and industry professionals to further the discussion on how to:
%   codify the principles from secure and privacy engineering into programming language constructs and or tools, 
%   create programming languages that have security and privacy as foundational tenets, and 
%   create/codify constructs or tools that enables secure and privacy-preserving (business) operations.
%   All submissions connected to the goals above, and their associated topics, are encouraged.
%%

%\subsection{Outline}
%
%An outline of this paper follows:
%
%\begin{itemize}
%  \item In \S 2, we define the type system which is embedded in Ace. We include a type safety proof for the string
%    segment of this language and prove the correctness of a translation to an underlying langguage $P$. In our theory,
%    $P$ is a simply typed lambda calculus equipped with a minimal regular expression library; in an implementation, $P$
%    stands in for Python or another underlying general-purpose programming language.
%  \item In \S 3, we discuss our implemention of this translation as a type system extension  within the Ace programming language.
%\end{itemize}

@@

\begin{figure}[t]
\begin{grammar}
<$r$> ::= $\epsilon$ | $.$ | $a$ | $r \cdot r$ | $r + r$ | $r*$ \hfill $a \in \Sigma$

@@

\begin{figure}[t]
\begin{grammar}

@@

\begin{figure}[t]\label{fig:lambdas}
\small
$\fbox{\inferrule{}{\sctx \vdash e : \sigma}}$
~~~~$\sctx ::= \emptyset \pipe \sctx, x : \sigma$
\begin{mathpar}
\inferrule[S-T-Var]
{ x:\sigma \in \sctx }
{ \sctx \vdash x:\sigma}

@@

\begin{figure}[t]
\small
$\fbox{\inferrule{}{e \sreduces e}}$
\begin{mathpar}
\inferrule[S-E-Abs]                                                             
{ \ }                                                                           
{\lambda x.e \sreduces \lambda x.e}

@@

A brief outline of this section follows:
\begin{itemize}
  \item Page 3 contains a definition of $\lambdas, \lambdap$ and the translation from $\lambdas$ to $\lambdap$. Grammar follows immediately at the top of page 4.
  \item In \S 2.1 we state some properties about regular expressions which are needed in our correctness proofs.
   \item In \S 2.2 we prove type safety for $\lambdap$ as well as both type safety and correctness for $\lambdas$.
  \item In \S 2.3 we prove that translation preserves the correctness result about $\lambdas$.
\end{itemize}

%The $\lambdas$ language is characterized by a type of strings indexed by regular
%expressions, together with operations on such strings which correspond to common
%input sanitation patterns.
%This section presents the grammar, typing rules and operational semantics for
%$\lambdas$ as well as an underlying language $\lambdap$.
%
%The system $\lambdas$ is the simply typed lambda calculus extended with \emph{regular expression types}, which are string
%types ensuring a string belongs to a specified language. For instance, 
%$S : \stringin{r}$ reads ``$s$ is a string matching $r$".
%the system includes an operation for replacing all instances of a pattern $r$ in a string $s_1$ with another string $s_2$. 
%Input sanitation algorithms -- as implemented by developers or within popular libraries and frameworks -- are often implemented in terms of this replace operation.
%
%The language $\lambdap$ is a simple functional language extended with a minimal regular expression library.
%Any general purpose programming language could stand in for $\lambdap$; for instance, SML or Python.
%In an implementation, our correctness results are modulo the underlying language's correct implementation of regular expression matching (see P-E-Replace).
%
%Finally, we define a translation from our type system $\lambdas$ into $\lambdap$ which preserves the safety guarnatee codified in the string types of $\lambdas$. Because our extension is conservative and its guarnatees are preserved under translation to an underlying language, it is highly composable with other analyses.
%
%Unfortunately, we are unable to present full proofs in this paper due to space constraints; however, proofs are given in full in the accompanying technical report?.
%%Note that we never insert an internal check where the type of the string implies
%%that a check must succeed. Furthermore, expensive calculations (such as language
%%inclusion or ${\sf rreplace}$) all occur at compile time.
%%Since sanitation problems are generally expressible with smalle, simple expressions,
%%we believe that the compile-time overhead is not significant enough to prohibiour target usecase. Informal experimentation with our implementation seems to
%%support this assertion.

% Begin Grammar 
% Removes the < > from grammar productions.

% <$\theta$> ::= $\theta \rightarrow \theta$ \hfill target types \alt
% $\str$ \alt $\regex$

% <P> ::= $\lambda x . e$ \hfill target terms \alt
%   $e e$ \alt
%   $\tstr{s}$ \alt
%   $\rx{r}$ \alt
%   $\tconcat{P}{P}$ \alt
%   $\preplace{P}{P}{P}$ \alt
%   $\tcheck{P}{P}$ 
%   \end{grammar}
% \caption{Syntax for the fragment of our target language, $\lambdap$, containing strings and statically constructed regular expressions.}
% \label{fig:lcsSyntax}
% \end{figure}

%The $\lcs$ language gives static semantics for common regular expression library
%functions. In this treatment, we include concatenation and filtering.
%The ${\tt filter}$ function removes all instances of a regular expression in a string,
%while concatenation ($+$) concatenates two strings.
%
%\subsection{Typing}
%
%The $\strin[r]$ type is parameterized by regular expressions; if $e:\strin[r]$,
%then $e \in r$. Mapping from an arbitrary $\str$ to a $\strin[r]$ requires
%defining an algorithm -- in terms of filter -- for converting a $\strin[.*]$
%into a $\strin[r]$. The static semantics of the language defines the types of
%operations on regular expressions in terms of well-understood properties about
%regular lanugages; we recall these properties in section 3.
%
%\subsection{Dynamics}
%
%There are two evaluation judgements: $e:T \reduces e'$ and $e:T \treduces i$.
%The $\reduces$ relation is between $\lcs$ expressions, while the $\treduces$
%relation is a mapping from $\lcs$ expressions into internal language expressions
%$i$ such that $i \ival$.
%
%Safety of the evaluation relation depends upon an injective mapping from $\lcs$ types info
%internal language types. This relation, $h$, is defined below.
%
%\subsection{Type Safety}
%
%The type safety proof relies upon some assumptions about the type system and
%dynamics of the internal language, as well as some properties of regular
%languages.
%
%There must exist a translation from $\lcs$ types to the types of the internal
%language. For the remainder of this paper, we call the type translation function
%$h$.
%
%\begin{defn}[Type Translation Function $h$]
%  The type translation function $h : Type \rightarrow IType$ is defined as follows:
%  \begin{itemize}
%    \item $\forall r. h(\strin[r]) = istr$
%    \item $h(\str) = istr$
%  \end{itemize}
%\end{defn}
%
%Additionally, we assume that the internal language contains an implementation of strings,
%together with operations for concatenation and filtering by regular expression.
%
%\begin{defn}[Types of internal values]
%  Let $\tstrf{s}$ range over string literals and $r$ over regular expressions.
%  Internal values are typed as follows:
%\begin{itemize}
%  \item If $e=\tstrf{s}$ then $e:istr$.
%  \item If $e=\ifilter{r}{\tstrf{s}}$ then $e:istr$.
%  \item If $e=\tstrf{s_1}+\tstrf{s_2}$ then $e:istr$.
%\end{itemize}
%\end{defn}
%
%For simplicity, we assume a fixed translation from
%$\lcs$ regular expressions to regular expressions recognizable by the internal
%language's regex library (in practice, a fixed translation is acceptable.)
%To summarize, we assume an internal language containing 
%a string type together with operations for string concatentation and filtering. We
%expect closure over strings for both operations.
%Finally, recall that ${\tt ifilter}$ is only needed for dynamic casts, which may be removed without
%descreasing the expressivity or even usability of the language.
%Finally, the semantics of the filter function are defined in terms of ${\tt rl\_filter}$,
%which is a static version of ${\tt ifilter}$.
%

% I can't achieve the correct latout here. We should have something like:
%       Lam_S Synax         Lam_P Syntax
%       Lam_S Sematics      LAm_P Semantics
%                Translation rules
% Or, alternatively, we can flip reflect things so that the entire page
% reads top-down (Lam_S syntax Lam_S smantics horizontally aligned w/ one
% another).

% Unfortunately, I can't get that to work. So, we're settling for something
% slightly ugly, where the fitures appear in the incorrect order and the 
% translation rules are on top.

% The translations rules *just won't* go to the bottom of the page, so 
% instead I am placing them at the top. Since the figures are out of order,
% it's even more important that we get the number gith. therefore, the counter
% is modified before and after this figure.

\begin{figure*}[t]\label{fig:trans}

@@

\inferrule[Tr-Abs?]
{ \sctx \vdash \trden{e} = \iota }
{\sctx \vdash \trden{ \lambda x.e } = \lambda x . \iota}

@@

\end{mathpar}
\caption{Translation from source terms ($e$) to target terms ($\iota$).
The translation is type-directed.}
\end{figure*}

\inferrule[P-T-Check]
{\tctx \vdash \iota_1 : \regex \\ \tctx, x:\str \vdash \iota_2 : \sigma \\ \tctx \vdash \iota_3 : \sigma}
{\tctx \vdash \tcheck{\iota_1}{x.\iota_2}{\iota_3} : \sigma}
\end{mathpar}
\caption{Typing rules for $\lambdap$.
The typing context $\tctx$ is standard.}
\label{fig:ityping}
\end{figure}

\inferrule[P-E-Check-OK]
{i \treduces \str{s} \\ s \in \lang{r}}
{\tcheck{r}{i}{x.i_1}{i_2} \treduces [\str{s}/x]i_1}

\inferrule[P-E-Check-NotOK]
{i \treduces \str{s} \\ s \not \in \lang{r}}
{\tcheck{r}{i}{x.i_1}{i_2} \treduces i_2}

{\iota_1 \treduces \rx{r} \\ \iota_2 \treduces \strin{s} \\ s \in \lang{r}}
{\tcheck{\iota_1}{\iota_2} \treduces \tstr{s}}

\end{mathpar}
\caption{Big step semantics for of $\lambdap$}.
\label{fig:ieval}
\end{figure}

The $\lambdas$ system extends the simply-typed lambda calculus with regular
expression types; an explanation of significant typing rule follows:
\begin{itemize}
  \item Rule S-T-Concat states that the type of concatenated strings is 
    obtained by concatenating the regular expressions for each string.
  \item Rule S-T-Replace defines the type of $\textsf{rreplace}$. The expression $\textsf{rreplace}$
    evaluates to the result of substituting every string matching $r$ in $s_1$
    with $s_2$. This operation corresponds to the \textsf{str_replace} function of PHP.
    The type of these expressions is defined in terms of an extra-linguistic \textsf{lreplace}
    function, which is defined later in this section.
    The expression $\lreplace{r}{r_1}{r_2}$ is obtained by
    starting with $r_1$, and replacing any subexpression matching $r$ with $r_2$.
  \item Rule S-T-SafeCoerce allows only safe coercions between string types by
    expoiting the decidability of language inclusion.
\end{itemize}

@@

% NRF commented out section 2 in order to determine the length of section 1.
% Replace the pattern ^% with an emptty string in the selection starting at
% the first empty line of section 2 and ending just before \end{document}.

\section{Implementation in {Atlang}}\label{atlang}
\begin{figure}\label{fig:example}
\begin{lstlisting}
@fn
def sanitize(s : string_in[r'.*']):
  return (s.replace(r'"', '&quot;') 
           .replace(r'<', '&lt;')
           .replace(r'>', '&gt;'))

@@

def main(db):
  input = sanitize(user_input())
  results = db.execute(results_query(input))
  return results_div(input) + format(results)
\end{lstlisting}
\vspace{-10px}
\caption{Regular string types in atlang, a library that enables static type checking for Python. }
\end{figure}\label{fig:impl}
\begin{figure}
\begin{lstlisting}
class string_in(atlang.Type):
  def __init__(self, rx):
    rx = rx_normalize(rx)
    atlang.Type.__init__(idx=rx)

@@

\begin{figure}
\begin{lstlisting}
output of successful compilation
\end{lstlisting}
\caption{Output of successful compilation.}
\end{figure}

\begin{figure}
\begin{lstlisting}
output of failed compilation
\end{lstlisting}
\caption{Output of failed compilation.}
\end{figure}

\subsection{An Example}

Figure \ref{fig:example} demosntrates the use of string types in atlang.
The sanitize function takes an arbitrary string and returns a string without
double quotes or left and right brackets. In this example, we use HTML
escape sequences.

>>>>

We will begin in Sec. \ref{calculus} by specifying this type system minimally, as a conservative extension of the simply typed lambda calculus called $\lambdas$. This allows us to specify the guarantees that the type system provides precisely. We also formally specify a translation from this calculus to a typed calculus with only standard strings and regular expressions, intending it as a guide to language implementors interested in building this feature into their own languages. This also demonstrates that no additional space overhead is required.

@@

@@

\begin{figure}[h]
\begin{grammar}
<$r$> ::= $\epsilon$ | $.$ | $a$ | $r \cdot r$ | $r + r$ | $r*$ \hfill $a \in \Sigma$

@@

@@

A brief outline of this section follows:
\begin{itemize}
  \item In \S 2.1-2.3, we define $\lambdas$, $\lambdap$ and the translation from one to the other.
  \item In \S 2.4 we state some properties about regular expressions which are needed in our correctness proofs.
   \item In \S 2.5 we prove type safety for $\lambdap$ as well as both type safety and correctness for $\lambdas$.
  \item In \S 2.6 we prove that translation preserves the correctness result about $\lambdas$.
\end{itemize}

The $\lambdas$ system extends the simply-typed lambda calculus with regular
expression types; an explanation of significant typing rule follows:
\begin{itemize}
  \item Rule S-T-Concat states that the type of concatenated strings is 
    obtained by concatenating the regular expressions for each string.
  \item Rule S-T-Replace defines the type of $\textsf{rreplace}$. The expression $\textsf{rreplace}$
    evaluates to the result of substituting every string matching $r$ in $s_1$
    with $s_2$. This operation corresponds to the \textsf{str_replace} function of PHP.
    The type of these expressions is defined in terms of an extra-linguistic \textsf{lreplace}
    function, which is defined later in this section.
    The expression $\lreplace{r}{r_1}{r_2}$ is obtained by
    starting with $r_1$, and replacing any subexpression matching $r$ with $r_2$.
  \item Rule S-T-SafeCoerce allows only safe coercions between string types by
    expoiting the decidability of language inclusion.
  \item The check rules allow casts between strings that cannot be guaranteed at compile time, but might pass a runtime check.
\end{itemize}

\inferrule[P-T-Check]
{\tctx \vdash \iota_r : \regex \\ \tctx \vdash \iota_1 : \str \\ \tctx, x:\str \vdash \iota_2 : \sigma \\ \tctx \vdash \iota_3 : \sigma}
{\tctx \vdash \tcheck{\iota_r}{\iota_1}{x . \iota_2}{\iota_3} : \sigma}
\end{mathpar}
\caption{Typing rules for $\lambdap$.
The typing context $\tctx$ is standard.}
\label{fig:ityping}
\end{figure}

\inferrule[P-E-Check-OK]
{\iota \treduces \tstr{s} \\ s \in \lang{r}}
{\tcheck{\rx{r}}{\iota}{x.\iota_1}{\iota_2} \treduces [\tstr{s}/x]\iota_1}

\inferrule[P-E-Check-NotOK]
{\iota \treduces \tstr{s} \\ s \not \in \lang{r}}
{\tcheck{\rx{r}}{\iota}{x.\iota_1}{\iota_2} \treduces \iota_2}
\end{mathpar}
\caption{Big step semantics for of $\lambdap$}.
\label{fig:ieval}
\end{figure}

\begin{figure}\label{fig:lambdas}
\small
$\fbox{\inferrule{}{\sctx \vdash e : \sigma}}$
~~~~$\sctx ::= \emptyset \pipe \sctx, x : \sigma$
\begin{mathpar}
\inferrule[S-T-Var]
{ x:\sigma \in \sctx }
{ \sctx \vdash x:\sigma}

@@

\begin{figure}
\small
$\fbox{\inferrule{}{e \sreduces e}}$
\begin{mathpar}
\inferrule[S-E-Abs]                                                             
{ \ }                                                                           
{\lambda x.e \sreduces \lambda x.e}

@@

\newpage
\ 
\newpage
\ 
\newpage

\begin{figure}[h]\label{fig:trans}

@@

\inferrule[Tr-Abs]
{ \sctx \vdash \trden{e} = \iota }
{\sctx \vdash \trden{ \lambda x.e } = \lambda x . \iota}

@@

\end{mathpar}
\caption{Translation from source terms ($e$) to target terms ($\iota$).
The translation is type-directed.}
\end{figure}

@@

\section{Implementation in {Atlang}}\label{atlang}
%\begin{figure}
%\begin{lstlisting}
%output of successful compilation
%\end{lstlisting}
%\caption{Output of successful compilation.}
%\begin{figure}
%\begin{lstlisting}
%output of failed compilation
%\end{lstlisting}
%\caption{Output of failed compilation.}

\subsection{An Example}\label{whatev}

\begin{figure}[h]\begin{lstlisting}
@fn
def sanitize(s : string_in[r'.*']):
  return (s.replace(r'"', '&quot;') 
           .replace(r'<', '&lt;')
           .replace(r'>', '&gt;'))

@@

def main(db):
  input = sanitize(user_input())
  results = db.execute(results_query(input))
  return results_div(input) + format(results)
\end{lstlisting}
\vspace{-10px}
\caption{Regular string types in atlang, a library that enables static type checking for Python. }   \label{examplefig}

\end{figure}

Figure \ref{examplefig} demonstrates the use of string types in atlang.
The sanitize function takes an arbitrary string and returns a string without
double quotes or left and right brackets. In this example, we use HTML
escape sequences.

\label{fig:impl}
\begin{figure}
\begin{lstlisting}
class string_in(atlang.Type):
  def __init__(self, rx):
    rx = rx_normalize(rx)
    atlang.Type.__init__(idx=rx)

@@
